---
title: "Data Preparation in R"
author: "Executed by Neha More"
date: "November 15, 2017"
output: html_document
---



---

###Aim:
####This Mini Project involves data preparation of dataset cencus_income.csv in order to make it fit for futher analysis and model building.

###Description:
1. Creating dummy variable for character variables.
2. Grouping similar category variables and making dummies.
3. Dealing with flag variables.(for numeric variables)
4. Converting the target Variable.(Y)

###Data Information:
census_income.csv is a csv file containing 32561 obs and 15 variables.It describes the income range of people with their characteristic attributes.The income range of people is >50k and <=50k which is stored in target variable Y.
We need to prepare data for the remaining (14) variables which can be further usefull in building models.


###Initial setup
loading library dplyr
```{r message=FALSE}
library(dplyr)
```
reading file census_income.csv into d
```{r}
d=read.csv("census_income.csv",stringsAsFactors = F)
View(d)
```
variables in d are as follows:
```{r}
names(d)
```
structure of dataset is:
```{r}
glimpse(d)
```
The dataset includes 9 character variables namely workclass, education, marital.status,  occupation, relationship, race, sex, native.country and Y(target variable) which needs to be converted accordingly in order to make it fit for modelling.

lets summarise the dataset
```{r}
summary(d)
```
finding out the number of distinct caztegories in each 8 character variables.(excludes Y as its the target variable)
```{r}
for(i in 1:ncol(d)){
  if(class(d[,i])=="character"){
    if(names(d)[i]!="Y"){
      message=paste("Number of categories in ",names(d)[i]," : ")
      num.cat=length(unique(d[,i]))
      print(paste0(message,num.cat))
    }
  }
}
```
###1. Creating dummy variable : (for variables having 6 or <6 categories:relationship,race,sex)
####creating dummy variable for race:
```{r}
table(d$race)

```
lets create 4 dummy variables considering others as base variable.
```{r}
d=d%>%
  mutate(race_AIE=as.numeric(race==" Amer-Indian-Eskimo"),
         race_API=as.numeric(race==" Asian-Pac-Islander"),
         race_Black=as.numeric(race==" Black"),
         race_White=as.numeric(race==" White")) %>% 
  select(-race)
```
####lets create dummy for sex(2category)
```{r}
table(d$sex)
nchar(d$sex[2]) #:there was space in the string of male and female.
```
```{r}
d=d %>% 
  mutate(sex_Female=as.numeric(sex==" Female"),
         sex_Male=as.numeric(sex==" Male")) %>% 
  select(-sex)
```
####lets create dummy for relationship(6category)
```{r}
table(d$relationship)

```
```{r}
d=d %>% 
  mutate(rel_h=as.numeric(relationship==" Husband"),
         rel_nif=as.numeric(relationship==" Not-in-family"),
         rel_oc=as.numeric(relationship==" Own-child"),
         rel_um=as.numeric(relationship==" Unmarried"),
         rel_w=as.numeric(relationship==" Wife")) %>%
  select(-relationship)
```
here other relative were taken as base variable and 5 dummies were made.

###2. Dummies for variables having similar categories.
####When categories present in a variable are too many,we group similar categories and then make dummies.
####combining similar category of variable workclass and making dummies:
```{r}
table(d$workclass)
```
here no of categories:9
dummy variables:8 but that too much hence we combine similar categories.
```{r}
round(prop.table(table(d$workclass,d$Y),1),1)
addmargins(round(prop.table(table(d$workclass,d$Y),1),1),2)
```
Here never worked and without pay is taken as base.and we create 5 dummies as follows:
```{r}
d=d %>%
  mutate(wc_1=as.numeric(workclass==" Self-emp-inc"),
         wc_2=as.numeric(workclass==" Federal-gov"),
         wc_3=as.numeric(workclass %in% c(" Local-gov"," Self-emp-not-inc"," State-gov")),
         wc_4=as.numeric(workclass==" Private"),
         wc_5=as.numeric(workclass==" ?")) %>%
  select(-workclass)
```
####combining similar category of variable education and making dummies:
```{r}
round(prop.table(table(d$education,d$Y),1),1)

```
here we create 6 dummies by combining similar categories.
```{r}
d=d %>%
  mutate(edu_1=as.numeric(education %in% c(" 10th"," 11th"," 12th"," 7th-8th"," 9th")),
         edu_2=as.numeric(education %in% c(" 1st-4th"," 5th-6th"," Preschool")),
         edu_3=as.numeric(education %in% c(" Assoc-acdm"," HS-grad"," Some-college")),
         edu_4=as.numeric(education ==" Assoc-voc"),
         edu_5=as.numeric(education==" Bachelors"),
         edu_6=as.numeric(education==" Masters")) %>%
  select(-education)
```

####combining similar category of variable marital status and making dummies:
```{r}
round(prop.table(table(d$marital.status,d$Y),1),1)

```

```{r}
d=d %>%
  mutate(ms_1=as.numeric(marital.status==" Never-married"),
         ms_2=as.numeric(marital.status %in% c(" Married-AF-spouse"," Married-civ-spouse"))) %>%
  select(-marital.status)
```

####combining similar category of variable occupation and making dummies:

```{r}
round(prop.table(table(d$occupation,d$Y),1),1)

```
```{r}
d=d %>%
  mutate(oc_1=as.numeric(occupation==" Exec-managerial"),
         oc_2=as.numeric(occupation==" Prof-specialty"),
         oc_3=as.numeric(occupation %in% c(" Protective-serv"," Sales"," Tech-support")),
         oc_4=as.numeric(occupation %in% c(" Craft-repair"," Transport-moving")),
         oc_5=as.numeric(occupation %in% c(" Priv-house-serv"," Other-service"))) %>%
  select(-occupation)
```

####combining similar category of variable native.country and making dummies:

```{r}
k=round(prop.table(table(d$native.country,d$Y),1),1)
k
```
```{r}
sort(k[,1]) #sorting makes it easier to combine
```


```{r}
d=d %>%
  mutate(nc_1=as.numeric(native.country %in% c(" Cambodia"," France"," India",
                                               " Iran"," Japan"," Taiwan"," Yugoslavia")),
         nc_2=as.numeric(native.country %in% c(" ?"," Canada"," China"," Cuba"," England",
                                               " Germany"," Greece"," Hong"," Italy",
                                               " Philippines")),
         nc_3=as.numeric(native.country %in% c(" Hungary"," Ireland"," Poland"," Scotland",
                                               " South"," Thailand"," United-States")),
         nc_4=as.numeric(native.country %in% c(" Columbia"," Dominican-Republic",
                                               " Guatemala"," Holand-Netherlands",
                                               " Outlying-US(Guam-USVI-etc)"))) %>%
  select(-native.country)
```
here 4 dummies are made for native.country.

####hence we succesfully converted all character variables into dummy.
```{r}
glimpse(d)
```
###3. Dealing with flag variables.
####used for non categorical data values(ie numeric data values)

####Dealing with variable capital gain
```{r}
table(d$capital.gain)

```
Here 0 occurs 29k times.
Hence lets find its percentage of occurance.
This will give % of observations where capital.gain is 0
```{r}
sum(d$capital.gain==0)/nrow(d)

```
More than 90% values are 0 , lets go ahead create a flag variable for this:
```{r}
d=d %>%
  mutate(cg_flag0=as.numeric(capital.gain==0))
```
hence wherever capital gain value is zero,it is mutated as 1,rest values take value 0.

####similarly lets do for variable capital loss
```{r}
table(d$capital.loss)

```
lets find its percenatge fo ocuurance
```{r}
sum(d$capital.loss==0)/nrow(d)

```
95% more is capital loss values equal to zero.hence lets flag it.
```{r}
d=d %>%
  mutate(cl_flag0=as.numeric(capital.loss==0))
```

###4. Converting the target Variable.(Y)
```{r}
d$Y=as.numeric(d$Y==" >50K")

```
All values where Y>=50 is made "1" else rest everything is made "0"

```{r}
glimpse(d)
```

###  *Data preparation is thus successfully done and can be further used for building models.*